---
title: "11: Generalized Linear Models (ANOVA)"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2020"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
2. Apply special cases of the GLM (ANOVA) to real datasets
3. Interpret and report the results of ANOVA in publication-style formats

## Set up
```{r, message = FALSE}
getwd()
library(tidyverse)
library(agricolae)

Litter <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")

# Set date to date format
Litter$collectDate <- as.Date(Litter$collectDate , format = "%Y-%m-%d")

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

### One-way ANOVA
A one-way ANOVA is the same test in practice as a two-sample t-test but for three or more groups. In R, we can  run the model with the function `lm` or `aov`, the latter of which which will allow us to run post-hoc tests to determine pairwise differences.

Does total litter biomass differ among sites at the Niwot Ridge LTER?
```{r}
# Wrangle the data
Litter.Totals <- Litter %>%
  group_by(plotID, collectDate, nlcdClass) %>%
  summarise(dryMass = sum(dryMass))
#USE THIS FOR MP!!! XX

# Test for normality. 
# Note: the important assumption for GLMs is normality of residuals, 
# not necessarily normality of raw data. See below when we plot the model.
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_040"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_041"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_046"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_047"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_051"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_057"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_058"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_061"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_062"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_063"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_064"])
shapiro.test(Litter.Totals$dryMass[Litter.Totals$plotID == "NIWO_067"])

qqnorm(Litter.Totals$dryMass); qqline(Litter.Totals$dryMass) #common in envt data to have big right tails

# Test for equal variance
# ANOVA is robust against departures from equal variance.
bartlett.test(Litter.Totals$dryMass ~ Litter.Totals$plotID)
#LOW P = variances are not close to each other 

# Format ANOVA as aov
Litter.Totals.anova <- aov(data = Litter.Totals, dryMass ~ plotID)
summary(Litter.Totals.anova)
#dry mass is y, plotid is x 
#sig pval so plot id is a sig predictor of drymass. 
# need a high residual df. 


# Format ANOVA as lm
Litter.Totals.anova2 <- lm(data = Litter.Totals, dryMass ~ plotID)
summary(Litter.Totals.anova2)
#y = alpha1+alpha2+alpha3+alpha4...


# Checking model fit and assumptions
# ANOVA is robust against departures from normality.
plot(Litter.Totals.anova2)

# Post-hoc test
TukeyHSD(Litter.Totals.anova) #only run on aov object, not lm object

# Extract groupings for pairwise relationships
Litter.Totals.groups <- HSD.test(Litter.Totals.anova, "plotID", group = TRUE)
Litter.Totals.groups #group= true shows groupings 
#if the letters match, they're not sig different 

# Graph the results
Litter.Totals.plot <- ggplot(Litter.Totals, aes(x = plotID, y = dryMass)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun.y = max, vjust = -1, size = 3.5,
               label = c("abc", "ab", "abcd", "bcd", "bcd", "a", 
                         "d", "abcd", "cd", "d", "abcd", "abcd")) +
  #stat_summary will place a variable in the plot... vjust adjusts height based on the last data point?? the labels here are associated with the plot ids and the post-hoc groupings...
  
  labs(x = "Plot ID", y = "Dry Mass (g)") +
  ylim(0, 85)
print(Litter.Totals.plot)
```

What information does the output give us? How might we report this information in a report?

> Total litter biomass differed significantly among sites at the Niwot Ridge LTE (Figre1; ANOVA F(11,198) = 4.813, p< 0.0001). #t-stat is the 4.813; the farther the t-stat is away from 0, the more likely there will be sig groupings in the dataset. 


### Non-parametric equivalent of ANOVA: Kruskal-Wallis Test
As with the Wilcoxon test, the Kruskal-Wallis test is the non-parametric counterpart to the one-way ANOVA. Here, the data from two or more independent samples are replaced with their ranks without regard to the grouping AND based on the between-group sum of squares calculations. 

For multiple comparisons, a p-value < 0.05 indicates that there is a significant difference between groups, but it does not indicate which groups, or in this case, months, differ from each other.

To analyze specific pairs in the data, you must use a *post hoc* test. These include the Dunn's test, a pairwise Mann-Whitney with the Bonferroni correction, or the Conover-Iman test.

```{r}
Litter.kw <- kruskal.test(Litter.Totals$dryMass ~ Litter.Totals$plotID)
Litter.kw

#sig p-val = sig difference between groups; report a chi square, df, pval 

# There are two functions to run the Dunn Test - equivalent of the post-hoc for kw. 
# dunn.test(Litter.Totals$dryMass, Litter.Totals$plotID, kw = T, 
#           table = F, list = T, method = "holm", altp = T)   #From package dunn.test
# dunnTest(Litter.Totals$dryMass, Litter.Totals$plotID)                    #From package FSA
```

## TWO-WAY ANOVA
### Main effects
A two-way ANOVA allows us to examine the effects of two categorical explanatory variables on a continuous response variable. What if we wanted to know how litter masses varied across functional groups and land cover class?

```{r}
# Format as aov
Litter.anova.2way <- aov(data = Litter, dryMass ~ functionalGroup + nlcdClass)
summary(Litter.anova.2way)
#predict drymass by functional group and nlcdclass (Separately)
#both are sig predictors 

# Format as lm
Litter.anova.2way2 <- lm(data = Litter, dryMass ~ functionalGroup + nlcdClass)
summary(Litter.anova.2way2)

summary(Litter$functionalGroup)
summary(Litter$nlcdClass)
#will show you rankings; the alphabetcal grouping is in order, so intercept is the first functional group and first nlcd...flowers in evergreen 
#ASK FOR HELP INTERPRETING!!! XX 

# Run a post-hoc test for pairwise differences
TukeyHSD(Litter.anova.2way)

```

### Interaction effects
We may expect the effects of functional group and land cover to be dependent on each other. For instance, we may expect needle biomass to be larger in evergreen forests than in other land types. In this case, we might choose to run an interaction effects two-way ANOVA, which will examine the individual effects of the explanatory variables as well as the interaction of the explanatory variables.

The output gives test statistics for each explanatory variable as well as the interaction effect of the explanatory variables. If the p-value for the interaction effect is less than 0.05, then we would consider the interaction among the explanatory variables to be significant.

```{r}
# Format as aov
Litter.anova.2way3 <- aov(data = Litter, dryMass ~ functionalGroup * nlcdClass)
summary(Litter.anova.2way3)
#EVERYTHING WAS SIG - interaction between all of them. 

# Format as lm
Litter.anova.2way4 <- lm(data = Litter, dryMass ~ functionalGroup * nlcdClass)
summary(Litter.anova.2way4)

```

If the interaction is significant, we interpret pairwise differences for the interaction. If the interaction is not significant, we interpret differences for the main effects only.
```{r}
# Run a post-hoc test for pairwise differences
TukeyHSD(Litter.anova.2way3)
#too much info and cuts it off, so instead do hsd test. 

Litter.interaction <- with(Litter, interaction(functionalGroup, nlcdClass))
Litter.anova.2way5 <- aov(data = Litter, dryMass ~ Litter.interaction)

Litter.groups <- HSD.test(Litter.anova.2way5, "Litter.interaction", group = TRUE)
Litter.groups
#you have to create an interaction object so that you can call it in the hsd test. 
# Plot the results
# Edit this graph to make it more visually appealing.
# Add letters to represent statistical groupings.
Litter.anova.plot <- ggplot(Litter, aes(y = dryMass, x = nlcdClass, color =  functionalGroup)) +
  geom_boxplot()+
stat_summary(geom = "text", fun.y = max, vjust = -1, size = 3.5,
               label = c("d", "d", "d", "a","d","d","d","d","d", "d", "d", "b","d","d","d","d","d", "d", "cd", "bc","d","d","d","d"), position = position_dodge(0.75))+
  
  labs(x = "Land Cover Class", y = "Dry Mass (g)") 
print(Litter.anova.plot)





Litter.Totals.plot <- ggplot(Litter.Totals, aes(x = plotID, y = dryMass)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  stat_summary(geom = "text", fun.y = max, vjust = -1, size = 3.5,
               label = c("abc", "ab", "abcd", "bcd", "bcd", "a", 
                         "d", "abcd", "cd", "d", "abcd", "abcd")) +
  
  labs(x = "Plot ID", y = "Dry Mass (g)") +
  ylim(0, 85)
print(Litter.Totals.plot)

```

