---
title: "12: Generalized Linear Models (Linear Regression)"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2020"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
2. Apply special cases of the GLM (linear regression) to real datasets
3. Interpret and report the results of linear regressions in publication-style formats
3. Apply model selection methods to choose model formulations

## Set up
```{r, message = FALSE}
getwd()
library(tidyverse)
options(scipen = 4) #gets rid of e^- and prints full #

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## Linear Regression
The linear regression, like the t-test and ANOVA, is a special case of the **generalized linear model** (GLM). A linear regression is comprised of a continuous response variable, plus a combination of 1+ continuous response variables (plus the error term). The deterministic portion of the equation describes the response variable as lying on a straight line, with an intercept and a slope term. The equation is thus a typical algebraic expression: 
#HOW TO MAKE EQUATIONS IN SCRIPT
$$ y = \alpha + \beta*x + \epsilon $$

The goal for the linear regression is to find a **line of best fit**, which is the line drawn through the bivariate space that minimizes the total distance of points from the line. This is also called a "least squares" regression. The remainder of the variance not explained by the model is called the **residual error.** 

The linear regression will test the null hypotheses that

1. The intercept (alpha) is equal to zero.
2. The slope (beta) is equal to zero

Whether or not we care about the result of each of these tested hypotheses will depend on our research question. Sometimes, the test for the intercept will be of interest, and sometimes it will not.

Important components of the linear regression are the correlation and the R-squared value. The **correlation** is a number between -1 and 1, describing the relationship between the variables. Correlations close to -1 represent strong negative correlations, correlations close to zero represent weak correlations, and correlations close to 1 represent strong positive correlations. The **R-squared value** is the correlation squared, becoming a number between 0 and 1. The R-squared value describes the percent of variance accounted for by the explanatory variables. 
#R is the correlation; -1 is strong neg; 1 is strong pos
#R squared removes the (-) component 

## Simple Linear Regression
For the NTL-LTER dataset, can we predict irradiance (light level) from depth?
```{r}
irradiance.regression <- lm(PeterPaul.chem.nutrients$irradianceWater ~ PeterPaul.chem.nutrients$depth)
# another way to format the lm function note you just call the dataframe once and then the columns. Preferred. 
irradiance.regression <- lm(data = PeterPaul.chem.nutrients, irradianceWater ~ depth)
summary(irradiance.regression)
#results: p-value, beta, alpha, adjusted r-sq (31% of variance in irradiance is explained by depth), df 

# Correlation: TELLS YOU R and (+/-)
cor.test(PeterPaul.chem.nutrients$irradianceWater, PeterPaul.chem.nutrients$depth)
#note the comma, not ~; R = -.55 R2 = .309
#
```
Question: How would you report the results of this test (overall findings and report of statistical output)?

>  Irradiance decreases significantly with increasing depth (linear regression, R2 = 0.31, df= 14,449, p < 0.0001). 

> For each 1m increase in depth, irradiance decreases by 95 units. 

So, we see there is a significant negative correlation between irradiance and depth (lower light levels at greater depths), and that this model explains about 31 % of the total variance in irradiance. Let's visualize this relationship and the model itself. 

An exploratory option to visualize the model fit is to use the function `plot`. This function will return four graphs, which are intended only for checking the fit of the model and not for communicating results. The plots that are returned are: 

1. **Residuals vs. Fitted.** The value predicted by the line of best fit is the fitted value, and the residual is the distance of that actual value from the predicted value. By definition, there will be a balance of positive and negative residuals. Watch for drastic asymmetry from side to side or a marked departure from zero for the red line - these are signs of a poor model fit.

2. **Normal Q-Q.** The points should fall close to the 1:1 line. We often see departures from 1:1 at the high and low ends of the dataset, which could be outliers. 

3. **Scale-Location.** Similar to the residuals vs. fitted graph, this will graph the squared standardized residuals by the fitted values. 

4. **Residuals vs. Leverage.** This graph will display potential outliers. The values that fall outside the dashed red lines (Cook's distance) are outliers for the model. Watch for drastic departures of the solid red line from horizontal - this is a sign of a poor model fit.

```{r, fig.height = 3, fig.width = 4}
par(mfrow = c(2,2), mar=c(1,1,1,1)) # mar allows you to print properly in knit
plot(irradiance.regression)
par(mfrow = c(1,1)) #reset each time!! 
#residual v fitted: clustering bad; outlier
#Normal q-q : outlier prob
#Scale-location: clustering again
#Cooks: outlier
```

The option best suited for communicating findings is to plot the explanatory and response variables as a scatterplot. 

```{r, fig.height = 3, fig.width = 4}
# Plot the regression
irradiancebydepth <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
  ylim(0, 2000) +
  geom_point() 
print(irradiancebydepth) 
#outlier: see that air irradiance is only 8000 so 25000 is not feasible. To remove it easily, set the ylim different.
#exponential decay in scatter. 
#consider logging to make it linear 
#remove outlier 
#you can't have zero values in logs; if there are only a few points, remove them or edit them to be non-zero. 
```

Given the distribution of irradiance values, we don't have a linear relationship between x and y in this case. Let's try log-transforming the irradiance values.

```{r, fig.height = 3, fig.width = 4}
PeterPaul.chem.nutrients <- filter(PeterPaul.chem.nutrients, 
                                   irradianceWater != 0 & irradianceWater < 5000) #remove the zeros and the outlier
irradiance.regression2 <- lm(data = PeterPaul.chem.nutrients, log(irradianceWater) ~ depth) #log irradiance by depth. 
summary(irradiance.regression2)
#logging changes the adjusted R2 TO OVER 70%; only use ajusted R2 when there are large samples. 

par(mfrow = c(2,2), mar=c(1,1,1,1))
plot(irradiance.regression2)
par(mfrow = c(1,1))

# Add a line and standard error for the linear regression
irradiancebydepth2 <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
  #geom_smooth(method = "lm") + #two different variables compared with a line of best fit; includes the 95% but you can barely see it because it's so well fit. Only include the line without the CI if you're not going to talk about CI OF THE MODEL. 
  geom_smooth(method = 'lm', se=FALSE, color = "black")+
  scale_y_log10() + #instead of doing y = log(irradianceWater); shows the scale as log of actual irradiance values instead of logged values - much easier to read. 
  geom_point(pch = 1, color = "blue")+ #designates the point type (hollow, color, etc)
  labs( x = "Depth (m)", y = "Irradiance (unk units)") 
          
print(irradiancebydepth2) 

#look up sentences to describe how to interpret logged values and results... 

# SE can also be removed
irradiancebydepth2 <- 
    ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
    geom_point() +
    scale_y_log10() +
    geom_smooth(method = 'lm', se = FALSE, color = "black")
print(irradiancebydepth2)

# Make the graph attractive

```

## Non-parametric equivalent: Spearman's Rho
As with the t-test and ANOVA, there is a nonparametric variant to the linear regression. The **Spearman's rho** test has the advantage of not depending on the normal distribution, but this test is not as robust as the linear regression.

``` {r}
cor.test(PeterPaul.chem.nutrients$irradianceWater, PeterPaul.chem.nutrients$depth, 
         method = "spearman", exact = FALSE)
```

## Multiple Regression
It is possible, and often useful, to consider multiple continuous explanatory variables at a time in a linear regression. For example, total phosphorus concentration in Paul Lake (the unfertilized lake) could be dependent on depth and dissolved oxygen concentration: 

``` {r, fig.height = 3, fig.width = 4}
TPregression <- lm(data = subset(PeterPaul.chem.nutrients, lakename == "Paul Lake"), 
                   tp_ug ~ depth + dissolvedOxygen)
summary(TPregression) # alpha = depth = 0 and do =0, beta1 = depth beta2 = do
#results: even with df of 288, you can use adjusted R2 (29%); 

TPplot <- ggplot(subset(PeterPaul.chem.nutrients, lakename == "Paul Lake"), 
                 aes(x = dissolvedOxygen, y = tp_ug, color = depth)) +
  geom_point() +
  xlim(0, 20)
print(TPplot) # not a good visual; hard to interpret 

#note the outliers and clustering; try transforming 
```

## Correlation Plots
We can also make exploratory plots of several continuous data points to determine possible relationships, as well as covariance among explanatory variables. 

```{r, fig.height = 3, fig.width = 4}
#install.packages("corrplot")
library(corrplot)
#test for multicollinearity 
#problems with NAs
PeterPaulnutrients <- 
  PeterPaul.chem.nutrients %>%
  select(tn_ug:po4) %>%
  na.omit()
PeterPaulCorr <- cor(PeterPaulnutrients) #create a matrix 
corrplot(PeterPaulCorr, method = "ellipse") #plot the matrix; stronger corr = line; weaker corr = circle, direction of elipse shows (+/-)
corrplot.mixed(PeterPaulCorr, upper = "ellipse")
```

## AIC to select variables

However, it is possible to over-parameterize a linear model. Adding additional explanatory variables takes away degrees of freedom, and if explanatory variables co-vary the interpretation can become overly complicated. Remember, an ideal statistical model balances simplicity and explanatory power! To help with this tradeoff, we can use the **Akaike's Information Criterion (AIC)** to compute a stepwise regression that either adds explanatory variables from the bottom up or removes explanatory variables from a full set of suggested options. The smaller the AIC value, the better. 

Let's say we want to know which explanatory variables will allow us to best predict total phosphorus concentrations. Potential explanatory variables from the dataset could include depth, dissolved oxygen, temperature, PAR, total N concentration, and phosphate concentration.

```{r}
Paul.naomit <- PeterPaul.chem.nutrients %>%
  filter(lakename == "Paul Lake") %>%
  na.omit()

TPAIC <- lm(data = Paul.naomit, tp_ug ~ depth + dissolvedOxygen + 
              temperature_C + tn_ug + po4)
step(TPAIC)
TPmodel <- lm(data = Paul.naomit, tp_ug ~ dissolvedOxygen + temperature_C + tn_ug)
summary(TPmodel)

#AIC runs until it has <none> on top, then it runs the lm. Pull that lm and make it an object. Note that tn was not sig, but it helps the AIC value (max R2 and max DF). 

```