---
title: "Assignment 2: Coding Basics"
author: "Kristine Swann"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A02_CodingBasics.Rmd") prior to submission.

The completed exercise is due on Tuesday, January 21 at 1:00 pm.


## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
funfunfun <- seq(1, 100, 4) #seq from to by, named

#2. 
funmean <- mean(funfunfun) #named mean to call it up
funmedian <- median(funfunfun) #same

#3. 
funmean > funmedian #test

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
names <- c("alex", "bea", "claire", "dan") # character/string vector
scores <- c(100,80,60,40) #numeric
pass <- c(TRUE,TRUE,TRUE,FALSE) #logical
dframe <-data.frame(names, scores, pass)
names(dframe)<-c("names", "scores", "pass"); View(dframe)
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: It's not a matrix...it's more like a csv...it doesn't have the same requirements as matrices would have for dimensions/characters, and you can't do linear algebra on it. 

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
if (pass ==TRUE) {print ("Next Level")} else {print("Complete Failure")}
ifelse(pass,"Next Level","Complete Failure")


```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:ifelse worked. Apparently, only ifelse works for logical vectors. But, I also tried using "scores" and >< values, but that came back with the same condition has length > 1 error.  
